---
title: "MockDilution"
author: "Erin Dahl"
date: "1/21/2020"
output: html_document
---

### Brief overview of the mock microbial dilution series data set

The data set generated for this manuscript is a dilution series of a mock microbial community. 
ZymoBIOMIC mock community standards was used, which consisted of eight bacterial species (*Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, *Bacillus subtilis*) and two fungal species (*Saccharomyces cerevisiae* and *Cryptococcus neoformans*). The mock microbial community was diluted with microbial free water (Qiagen) in eight rounds of a serial three-fold dilution prior to DNA extraction. 

The V4 region of the 16S rRNA gene was amplified and sequenced with Illumina MiSeq. The raw sequencing data is available in the Sequence Read Archive (SRA) under accession number [SRP155048](https://www.ncbi.nlm.nih.gov/sra/SRP155048). Reads were processed using DADA2 to generate amplicon sequence variants (ASVs), which were used to generate the phyloseq object used in this analysis. 

## Initial analysis of the mock community dilution series data set

### Set up the workspace

```{r set workspace, message=FALSE}
# load libraries

library(phyloseq)
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(reshape2)
library(knitr)
options(digits=2) #sig figs
library(dplyr)
library(ggpubr)
library(cowplot)

#save session info (packages and versions loaded)

```

### Create functions to use in this analysis

We created a function called expCompBarPlot to display the expected mock community ASVs in color and unexpected contaminants sequences in grayscale. 


```{r defineFunctions}
# Create function to bar plots with contaminants in grey scale and expected mock community sequences in color

expCompBarPlot <- function(physeq, exp_taxa, title){
  # physeq - phyloseq object that will be plotted
  # exp_taxa - taxa that are expected to be in the mock community
  # title - title for plot
  
# set up data table
data_table <- as.data.frame(t(physeq@otu_table)) 
#phylogenetic sequencing experiment as an instance of the phyloseq class and OTU table

data_table$reference = FALSE
  data_table$reference[rownames(data_table) %in% exp_taxa] = TRUE
  sample_names <- sample_names(physeq)
  data_table$id <- paste0('ASV_', 1:nrow(data_table))
  dilution_labels <- sample_data(physeq)$Dilutions

  set.seed(444)

# define the colors to use for reference and non-reference OTUs/ASVs
  ref_colors <- brewer.pal(sum(data_table$reference), "Paired")
  other_colors <- sample(grey.colors(5, start = 0.5, end = 0.9), sum(!data_table$reference), replace = TRUE)

# add a color variable to the data table
  data_table$color <- rep(NA, nrow(data_table))
  data_table$color[data_table$reference] <- ref_colors
  data_table$color[!data_table$reference] <- other_colors


# reshape the data table into a ggplot-friendly format, by gathering samples into a single column called "count"

color_gg <- data_table %>% select(id, sample_names, color) %>% gather("sample", "count", sample_names)
legend_color <- c(bright = ref_colors[2], dull = other_colors[2])
data_gg <- data_table %>% gather("sample", "count", sample_names)

 data_gg <- inner_join(data_gg,color_gg)

# create the composition bar plot
# geom_col makes the heights of the bars represent values in the data
  comp_bar <- ggplot(data_gg, aes(x = sample, y = count)) +
    geom_col(aes(fill = color, group = reference, alpha = ifelse(reference, "bright", "dull")), width = 0.7, position = position_fill()) +
    scale_fill_identity(guide = FALSE) +
    scale_alpha_manual(name = "Sequence type",
                     labels = c("expected sequences", "other"),
                     values = c(bright = 1, dull = 1),
                     guide = guide_legend(override.aes = list(fill = c(ref_colors[4], "#AEAEAE")),
                                          keywidth = NULL, keyheight = NULL)) +
    labs(title = title, x = "sample", y = "Relative Abundance") +
    theme(legend.position = "right", legend.title = element_text(size = 12),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))
  comp_bar
}

```

### Load the Data

Make sure RDataFile and RMarkdownfile save in same dir

```{r loadData}
## Load the dataset
load("mockDilutions.RData")
```

This data set includes 3 phyloseq objects:

* ps - phyloseq object with the mock community dilution series and blank control
* mock_ps - phyloseq object with the mock community dilution series only
* blank_ps - phyloseq object with the blank control only

and one other object:
* asv_key - a matrix containing ASV name : ASV sequence pairs 

mock_ps and blank_ps were created from the ps phyloseq object by subsetting the desired samples to make the summaries of these samples more straight forward. The asv_key object was created to make downstream plotting and summarizing easier while being able to retain the full asv sequences for each asv. 

### Evaluating the undiluted mock community sample

We expect the undiluted mock community sample to mostly contain the expected sequences from the mock community. To evaluate this, we subset out the undiluted mock community sample (D0) as a new separate phyloseq object and remove the ASVs that are not present in the sample. Since the mock community contains eight bacteria, we expect there to be eight ASVs.

```{r }

# Create profile of only expected sequences from the undiluted mock community sample

# Subset the undiluted mock microbial sample  (sample name 'D0')
mock_ps_pure <- subset_samples(mock_ps, sample_names(mock_ps)=='D0')

# Remove ASV's that are not present in the undiluted sample
mock_ps_pure <- prune_taxa(taxa_sums(mock_ps_pure)>0, mock_ps_pure)

# Change the SampleType and sample_names of the pure mock microbial community sample 

sample_data(mock_ps_pure)$SampleType <- 'MockCommunityProfile'
sample_names(mock_ps_pure) <- paste ('mc', sample_names(mock_ps_pure), sep = '_')

# display a summary of the new phyloseq object
mock_ps_pure

```

The phyloseq object for the undiluted mock community contains 18 ASVs instead of the expected eight ASVs. Further inspection of these ASVs indicates the eight expected sequences are present, and that there are 10 additional unexpected ASVs. One ASV is one nucleotide off from an expected sequence and is present in high abundance. This likely represents an actual mutation that occurred in our mock community. The other nine sequences are present in low abundance (< 60 reads out of a total of 222,159 reads) and are distinct from the expected sequences. These are low level contaminants which may be due to sample ‘cross talk’ - errors that occur from barcode switching during sequencing. We will remove these ASVs from the mock_ps_pure phyloseq object.

```{r}
# remove the unexpected ASVs from the undiluted mock microbial community dilution series

# make a list of the top 9 abundant ASV taxa names (this is plausible for filtering since the 9 sequences we want to remove are present in low abundance)

mock_taxa = names(sort(taxa_sums(mock_ps_pure), decreasing = TRUE)[1:9])

# subset the taxa in mock_ps_pure so only the expected sequences are present

mock_ps_pure <- prune_taxa(mock_taxa, mock_ps_pure)
# removes unwanted OTUs

```

### Identifying the impact of contaminants with decreasing starting material
Now that we know what the expected sequences of our mock microbial community are in our data set, we evaluate the unexpected sequences across the dilution series. 

```{r }
# display a summary of the mock community dilution series phyloseq object
mock_ps
```

A brief overview of the phyloseq object indicates that there are 755 ASVs in the mock microbial dilution series- that is a LOT more than the expected nine ASVs! 

Now we identify how the decreasing starting material impacts the number of reads per sample and the proportion of reads per sample that are unexpected contaminant sequences.  

```{r fig.width=5, fig.height=4}

# create a phyloseq object that is normalized to 100 (relative abundance)
ps_norm <- transform_sample_counts(ps,function(x) 100* x/sum(x))
mock_ps_norm <- transform_sample_counts(mock_ps,function(x) 100* x/sum(x))

# Identify the proportion of each sample that is the expected mock community ASVs
ps_norm_exp <- prune_taxa(mock_taxa,ps_norm)

# Create a table with the dilution, number of reads per sample, and proportion of contaminants per sample
dilutionSummary <- data.frame(DilutionSeries = sample_names(ps),NumberOfReads = sample_sums(ps), PercentContaminants = 100-sample_sums(ps_norm_exp))

# Create a variable to indicate the sample order of the plots
dilutions<-c('D0','D1','D2','D3','D4','D5','D6','D7','D8', 'Blank')

# Create plots to summarize these data
## Plot Figure 1A - number of reads per sample across dilution series
dilutionSummary %>% subset(DilutionSeries %in% dilutions[1:9]) %>%
  ggplot(.,aes(x=DilutionSeries, y= NumberOfReads)) + geom_bar (stat="identity", fill="steelblue") +
  theme_minimal() +scale_x_discrete(limits = dilutions) +
  labs (x= "Dilution Series", y= "Number of Reads") +
  theme (axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))
```

```{r Percent Contaminants Graph}
## Plot Figure 1B - Percent of contaminants across dilution series
dilutionSummary %>% subset(DilutionSeries %in% dilutions[1:9] ) %>%
  ggplot(., aes(x = DilutionSeries, y = PercentContaminants)) + geom_point(size = 3) +
  scale_x_discrete(limits = dilutions) +   
  labs(x = "Dilution Series", y = "Percent Contaminants") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))
```

```{r Stacked Bar Plot}
expCompBarPlot(ps_norm, mock_taxa, 'Initial Mock Microbial Community Dilution') + scale_x_discrete(limits = dilutions)
# scale_x_discrete() allows for labels between the bars in the bar chart

```

```{r #Sequences per Sample}
## Table 1 
# Number of sequences per sample

sample_sums (mock_ps)
```

```{r #ASVs per Sample}
# number of ASVs per sample

temp <- as.data.frame(ps@otu_table)
temp[temp > 0] <- 1
rowSums(temp)
```


###Evaluate the blank control

```{r}
blank_ps_norm <- subset_samples(ps_norm,sample_names(ps_norm) %in% c('Blank'))
blank_ps_norm <- prune_taxa(taxa_sums(blank_ps_norm) > 0, blank_ps_norm)
blank_ps_norm
```

```{r}
# 655 unique ASVs

#identify ASVs from mock community present in blank
table(taxa_names(blank_ps_norm) %in% mock_taxa)

```

```{r}
# Get genera present with >5% abundance
blank_ps_norm %>% subset_taxa(taxa_names(blank_ps_norm) %in% taxa_names(mock_ps_pure)) %>% psmelt() %>% select(Abundance, Family, Genus)

```

```{r}
# Collapse at the genus level, keeping unassigned genera

blank_ps_norm <- tax_glom(blank_ps_norm, "Genus", NArm = FALSE)
blank_ps_norm

```

```{r}
blank_ps_norm_melt <- psmelt(blank_ps_norm)
# Get genera present with >5% abundance
blank_ps_norm_melt %>% filter(Abundance > 5) %>% select(Abundance, Family, Genus)
```

```{r}
# Get genera present with abundance between 1 and 5%
blank_ps_norm_melt %>% filter(Abundance <= 5) %>% filter(Abundance > 1) %>% select(Abundance, Family, Genus)
```

### Evaluating contaminant ASVs

To evaluate the contaminant ASVs, we create a phyloseq object only containing the contaminant ASVs.

```{r}
# create a list of unexpected sequences (contaminants)

# create a list of all ASV taxa names
contaminant_taxa<-taxa_names(mock_ps)
# remove the expected mock community ASV taxa names
contaminant_taxa <- contaminant_taxa[!(contaminant_taxa %in% mock_taxa)]

# create a phyloseq object that only contains the contaminant sequences (for use with sourcetracker)
contaminants_ps<-prune_taxa(contaminant_taxa,mock_ps)
contaminants_ps<- prune_taxa(taxa_sums(contaminants_ps)>0,contaminants_ps)

# change the sample names to indicate that these samples only contain contmaminant ASVs
sample_names(contaminants_ps)<-paste('con',sample_names(contaminants_ps),sep = '_')
sample_data(contaminants_ps)$SampleType<-'ContaminantProfile'

# create phyloseq object from normalized data to summarize contamiant contribution
contaminants_ps_norm<-prune_taxa(contaminant_taxa,mock_ps_norm)
contaminants_ps_norm<- prune_taxa(taxa_sums(contaminants_ps_norm)>0,contaminants_ps_norm)

```

Now, we use the contaminants_ps object to evaluate the number contaminant ASVs that are present in the blank control sample.

```{r}
# Number of contaminant ASVs across dilution series
print(paste('Total number of contaminant ASVs', length(taxa_names(contaminants_ps))))

```

```{r}
# abundance of contaminant ASVs across samples
sample_sums(contaminants_ps_norm)
```

```{r}
# Abundance of contaminant genera diltuion series
contaminant_genera <- tax_glom(contaminants_ps_norm, 'Genus', NArm = FALSE) %>% 
  psmelt() 

# Maximum abundance of contaminant genera per diltuion sample
contaminant_genera %>%
  group_by(Sample) %>%
  filter(Abundance == max(Abundance)) %>% 
  select(Sample, Abundance)
```


```{r}
# Create list of contamiant ASVs with abundance > 1% (Supplemental Table 2)
contam_table <-contaminants_ps_norm %>% 
  psmelt() %>%
  group_by(OTU) %>%
  mutate(max_abundance = max(Abundance)) %>% 
  filter(max_abundance > 1) %>%
  mutate(in_n_samples = sum(Abundance > 0)) %>%
  mutate(in_blank = OTU %in% taxa_names(blank_ps_norm)) %>%
  mutate(asv_sequence = asv_key[asv_key$asv_name == OTU,]$asv_sequence) %>%
  select( Phylum, Class, Order, Family, Genus,max_abundance, in_n_samples, in_blank, asv_sequence) %>% 
  unique()
```

```{r}
# number of ASVs present in blank sample
print(paste('Number of contaminant ASVs also present in blank', length(intersect(taxa_names(contaminants_ps),taxa_names(blank_ps)))))
```

```{r}
# create a list of contaminants taxa that are not present in the blank control
contaminant_taxa_no_blank<-taxa_names(contaminants_ps)
contaminant_taxa_no_blank <- contaminant_taxa_no_blank[!(contaminant_taxa_no_blank %in% taxa_names(blank_ps))]

# Create  a binary list of contaminant ASVs indicating if the ASV is present in the blank control (1) or not (0)
contaminants_in_blank <- data.frame(matrix(1, ncol = length(taxa_names(contaminants_ps)), nrow = 1))
colnames(contaminants_in_blank) <- taxa_names(contaminants_ps)
contaminants_in_blank[,contaminant_taxa_no_blank] <- 0 
contaminants_in_blank <- t(contaminants_in_blank)
```

We found that there are a lot of contaminant ASVs present in the dilution series that aren’t in the blank control sample.

We also determine the proportion of contaminant ASVs that are not in found in the blank control to help us identify the actual impact of these.

```{r}
# Identify the contribution per sample of contaminants that are not present in blanks
# generate a phyloseq object with contaminants only normalized to 100 
contaminant_ps_norm <- transform_sample_counts(contaminants_ps,function(x) 100* x/sum(x))
contaminant_no_blanks<-prune_taxa(contaminant_taxa_no_blank,contaminant_ps_norm)

# Plot the proportion of contaminant ASVs per sample that were not present in the blank control
# Original plot; too many black lines, unable to see all different genus present
plot_bar(contaminant_no_blanks,fill='Genus', title = ' Proportion of Contaminant ASVs Not in the Blank Control Sample') + theme(legend.position='none') + ylim(c(0,100))

# Remove black lines from original plot
plot_bar(contaminant_no_blanks,fill='Genus', title = ' Proportion of Contaminant ASVs Not in the Blank Control Sample') + theme(legend.position='none') + ylim(c(0,100)) + geom_bar(stat="identity")

# Explore filling at kingdom level, may be more informative/may give a broader view
plot_bar(contaminant_no_blanks,fill='Kingdom', title = ' Proportion of Contaminant ASVs Not in the Blank Control Sample') + theme(legend.position='none') + ylim(c(0,100))

# Filling at a Kingdom level, all contaminants are from bacteria kingdom
plot_bar(contaminant_no_blanks,fill='Kingdom', title = ' Proportion of Contaminant ASVs Not in the Blank Control Sample') + ylim(c(0,100)) + geom_bar(stat="identity")

# Allows for scaling the legend to fit all varibles
leg_size = 0.3

# Explore filling at Phylum level, no black lines, includes phylum legend
# This fill method seems to be the most informative, also the legend is scaled
plot_bar(contaminant_no_blanks,fill='Phylum', title = ' Proportion of Contaminant ASVs Not in the Blank Control Sample') + geom_bar(stat="identity") + ylim(c(0,100)) + theme(legend.key.size = unit(leg_size, "cm"))

# Explores filling at Phylum level, uses a Blue Palette, only issue is that there are not enough unique colors in the Palette, so orig coloring may be best for this
plot_bar(contaminant_no_blanks,fill='Phylum', title = ' Proportion of Contaminant ASVs Not in the Blank Control Sample') + geom_bar(stat="identity") + ylim(c(0,100)) + scale_fill_brewer(palette = "Blues") + theme(legend.key.size = unit(leg_size, "cm"))

# Explores filling at Phylum level, uses default Palette
plot_bar(contaminant_no_blanks,fill='Phylum', title = ' Proportion of Contaminant ASVs Not in the Blank Control Sample') + geom_bar(stat="identity") + ylim(c(0,100)) + theme(legend.key.size = unit(leg_size, "cm"))
```


```{r}
sample_sums(contaminant_no_blanks)
```


```{r}
# sum and summarize the amount of contaminant signal arising from ASVs in blank control
100 - sample_sums(contaminant_no_blanks)
```

```{r}
summary(100 - sample_sums(contaminant_no_blanks))
```

From this plot, it is evident that the contaminants not present in the blank control make up a significant proportion of the contaminants. This is a bit surprising, and may be due to the blank control not representing all contaminants. In this experiment all blank controls were pooled into a single sample- it likely would have been better to not pool the blank controls.

We also evaluated how many times each contaminant ASV is present across the dilution series.

```{r}
# Count number of contaminants present in only one sample
contaminant_bin<-as.data.frame(contaminants_ps@otu_table)
contaminant_bin[contaminant_bin>0]<-1
contaminant_bin = t(contaminant_bin)
contaminant_nsamples <- rowSums(contaminant_bin)
table(contaminant_nsamples)

```

```{r}
# summarize information about contaminant ASVs only in one sample
contaminant_one_sample <- rownames(contaminant_bin[rowSums(contaminant_bin)==1,])
temp <- contaminants_ps %>% prune_taxa(contaminant_one_sample,.) %>% taxa_sums(.)
# min amount of contaminant ASVs in one sample
min(temp)
```

```{r}
# max amount of contaminant ASVs in one sample
max(temp)
```

```{r}
# average amount of contaminant ASVs in one sample
mean(temp)
```

### Impact of Contaminants on Alpha Diversity

Next, we evaluate the impact that the contaminant ASVs have on commonly used alpha diversity metrics.

```{r}
mock_alpha <- estimate_richness(mock_ps, measures=c("Observed", "Shannon", "InvSimpson"))
mock_alpha_pure <- estimate_richness(mock_ps_pure,  measures=c("Observed", "Shannon", "InvSimpson"))
plot_richness(mock_ps, measure = c('Observed','Shannon', 'InvSimpson') )
```

```{r}
# calculate the difference between the observed and actual alpha diversity measures 
max_diff <- ((mock_alpha[9,] - mock_alpha_pure) / mock_alpha_pure) 
```

Finally, we save the work space image so that we can revisit the analysis if needed and so that we can easily load the processed data for the next steps of analysis.

```{r}
# Save workspace
save.image("mockDilutionsPrep.RData")
```